---
params:
  echoCode: TRUE
  mainTitle: "`ggplot2` Solutions: Gapminder Data"
title: '`r params$mainTitle`'
author: "Type Your Name Here"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$echoCode, 
                      eval = TRUE, 
                      cache = FALSE, 
                      message = FALSE)
options(width=100)
```

# Gapminder Data

We will use a subset of the gapminder data[^gapminder]. The variables in the 
dataset are as follows.

- `continent`: Continent
- `country`: Country name
- `year`: Year
- `pop`: Population
- `gdpPercap`: GDP per capita (US$, inflation-adjusted)
- `lefeexp`: Life expectancy at birth, in years
- `infmort`: Infant Mortality (rate per 1000 births)
- `fertility`: Children per woman

## Read in the data

```{r importgmshow, echo = TRUE}
library(data.table)
library(here)

gm <- fread(here("data", "gapminder.csv"))

# Some modifications which will be useful later
gm[, continent:=as.factor(continent)]
gm[, country:=as.factor(country)]
gm[, fyear:=as.factor(year)]
gm
str(gm)

# A dataset with just Portuguese data
gmPT <- gm[country == "Portugal"]
gmPT
```

# Life Expectancy - Portugal

Let's plot life expectancy over time for Portugal.
```{r lifeexpP, echo = TRUE}
library(ggplot2)
library(magrittr)

gm[country == "Portugal"] %>% 
ggplot(aes(year, lifeexp)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous(limits=c(0, 80)) 
```


## Exercise 1: Portugal

### Infant Mortality - Portugal

Plot the infant mortality rate over time for Portugal. Include the following:

+ Colour for the lines and points.
+ Labels for the axes.
+ Add "Infant Mortality - Portugal" as a title (hint: use `ggtitle()`).

```{r infmortPT-1}
gm[country == "Portugal"] %>% 
ggplot(aes(year, infmort)) +
  geom_point(colour="blue") + 
  geom_line(colour="blue") +
  scale_y_continuous(limits=c(0, 100)) +
  ggtitle("Portugal - Infant Mortality") +
  xlab("Year") +
  ylab("Infant Mortality (rate per 1000 births)")

```

**Interpret this plot.**

### Fertility Rate - Portugal

Plot the fertility rate over time for Portugal. Include the following:

+ Colour for the lines and points.
+ Labels for the axes.
+ Add "Fertility Rate - Portugal" as a title (hint: use `ggtitle()`).

```{r fertilityPT-1}
gm[country == "Portugal"] %>% 
ggplot(aes(year, fertility)) +
  geom_point(colour="darkred") + 
  geom_line(colour="darkgreen") +
  scale_y_continuous(limits=c(0, 4)) +
  ggtitle("Portugal - Fertility Rate") +
  xlab("Year") +
  ylab("Children per woman")

```

Use the colours of the Portuguese flag,: "darkred" for points and "darkgreen" 
for lines. What do you notice when you switch the order of `geom_point()` and
`geom_line()`? Do you strictly need to include the `geom_point()`?

```{r fertilityPT-2}
gm[country == "Portugal"] %>% 
ggplot(aes(year, fertility)) +
  geom_line(colour="darkgreen") +
  geom_point(colour="darkred") + 
  scale_y_continuous(limits=c(0, 4)) +
  ggtitle("Portugal - Fertility Rate") +
  xlab("Year") +
  ylab("Children per woman")
```

**Interpret this plot.**

### Population & GDP

Produce the same plots as above for population and GDP per capita. How do you 
interpret them?

```{r popPT}
gm[country == "Portugal"] %>% 
ggplot(aes(year, pop)) +
  geom_point(colour="blue") + 
  geom_line(colour="blue") +
  scale_y_continuous(limits=c(0, 11000000)) +
  ggtitle("Portugal - Total Population") +
  xlab("Year") +
  ylab("Total Population")
```

```{r gdpPercapPT}
gm[country == "Portugal"] %>% 
ggplot(aes(year, gdpPercap)) +
  geom_point(colour="blue") + 
  geom_line(colour="blue") +
  scale_y_continuous(limits=c(0, 12500)) +
  ggtitle("Portugal - GDP per capita") +
  xlab("Year") +
  ylab("GDP per capita (US$)")
```


# Life Expectancy - Europe

## Exercise 2: Europe

In this exercise we will evaluate how Portugal performs in Europe for:

- Infant Mortality
- Fertility Rate
- GDP per Capita

Produce the follow plots:

- Boxplot over time
- Line plot over time with European countries in grey and Portugal in blue.

### Infant Mortality - Portugal in Europe

The performance of Portugal in Europe for infant mortality rate (per 1000 births).

```{r infmort-boxplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(fyear, infmort)) + 
  geom_boxplot(outlier.size = .5) + 
  geom_point(data=gmPT, shape = 4, colour="blue") + 
  scale_x_discrete(breaks=seq(1960, 2010, 10)) + 
  scale_y_continuous(limits=c(0, 125)) +
  ggtitle("Portugal in Europe - Infant Mortality") +
  xlab("Year") +
  ylab("Infant Mortality (rate per 1000 births)")
```

```{r infmort-lineplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(year, infmort, group=country)) + 
  geom_line(colour="grey80") +
  geom_line(data=gmPT, colour="blue") + 
  scale_y_continuous(limits=c(0, 125)) +
  ggtitle("Portugal in Europe - Infant Mortality") +
  xlab("Year") +
  ylab("Infant Mortality (rate per 1000 births)")
```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

### Fertility Rate - Portugal in Europe

The performance of Portugal in Europe for the total fertility rate (children per woman).

```{r fertility-boxplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(fyear, fertility)) + 
  geom_boxplot(outlier.size = .5) + 
  geom_point(data=gmPT, shape = 4, colour="blue") + 
  scale_x_discrete(breaks=seq(1960, 2010, 10)) + 
  scale_y_continuous(limits=c(0, 7)) +
  ggtitle("Portugal in Europe - Fertility Rate") +
  xlab("Year") +
  ylab("Children per woman")
```

```{r fertility-lineplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(year, fertility, group=country)) + 
  geom_line(colour="grey80") +
  geom_line(data=gmPT, colour="blue") + 
  scale_y_continuous(limits=c(0, 7)) +
  ggtitle("Portugal in Europe - Fertility Rate") +
  xlab("Year") +
  ylab("Children per woman")
```


**Interpret this plot.** How does Portugal perform relative to other EU countries?

### GDP - Portugal in Europe

The performance of Portugal in Europe for the GDP per capita (inflation adjusted US$).

```{r GDP-boxplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(fyear, gdpPercap)) + 
  geom_boxplot(outlier.size = .5) + 
  geom_point(data=gmPT, shape = 4, colour="blue") + 
  scale_x_discrete(breaks=seq(1960, 2010, 10)) + 
  scale_y_continuous(limits=c(0, 50000)) +
  ggtitle("Portugal in Europe - GDP per Capita") +
  xlab("Year") +
  ylab("GDP per Capita (US$)")
```

```{r GDP-lineplot}
gm[continent=="Europe"] %>% 
  ggplot(aes(year, gdpPercap, group=country)) + 
  geom_line(colour="grey80") +
  geom_line(data=gmPT, colour="blue") + 
  scale_y_continuous(limits=c(0, 50000)) +
  ggtitle("Portugal in Europe - GDP per Capita") +
  xlab("Year") +
  ylab("GDP per Capita (US$)")
```


**Interpret this plot.** How does Portugal perform relative to other EU countries?

# Life Expectancy - Europe

## Exercise 3: Gapminder plot

Reproduce the "Health versus Wealth" gapminder plot but replacing health with:

- Infant Mortality
- Fertility Rate

### Gapminder Plot - Infant Mortality

```{r infmort-gdp-gapminder}
yrs <- c(1960, 1980, 2000, 2010)
gm[continent=="Europe" & year %in% yrs] %>% 
  ggplot(aes(gdpPercap, infmort)) + 
  geom_point(aes(size=pop, colour=fyear), shape = 21) +
  scale_x_log10(
    breaks = c(10^3, 10^4, 10^5),
    labels = c("1k","10k","100k")) + 
  scale_y_continuous(limits=c(0, 125)) +
  ggtitle("Portugal - Health vs Wealth") +
  xlab("GDP per Capita (US$)") +
  ylab("Infant Mortality (rate per 1000 births)") +
  facet_wrap(~fyear) +
  geom_point(data = gmPT[year %in% yrs], colour="blue")
```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

### Gapminder Plot - Fertility


```{r fertility-gdp-gapminder}
yrs <- c(1960, 1980, 2000, 2010)
gm[continent=="Europe" & year %in% yrs] %>% 
  ggplot(aes(gdpPercap, fertility)) + 
  geom_point(aes(size=pop, colour=fyear), shape = 21) +
  scale_x_log10(
    breaks = c(10^3, 10^4, 10^5),
    labels = c("1k","10k","100k")) + 
  scale_y_continuous(limits=c(0, 7)) +
  ggtitle("Portugal - Health vs Wealth") +
  xlab("GDP per Capita (US$)") +
  ylab("Children per woman") +
  facet_wrap(~fyear) +
  geom_point(data = gmPT[year %in% yrs], colour="blue")
```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

[^gapminder]: Source: https://www.gapminder.org/data/
