---
title: '`ggplot2` Exercises: Gapminder Data'
author: "Put Your Name Here"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
options(width=100)
```

# Gapminder Data

We will use a subset of the gapminder data[^gapminder]. The variables in the 
dataset are as follows.

- `continent`: Continent
- `country`: Country name
- `year`: Year
- `population`: Population
- `gdp`: GDP per capita (US$, inflation-adjusted)
- `lefeexp`: Life expectancy at birth, in years
- `infmort`: Infant Mortality (rate per 1000 births)
- `fertility`: Children per woman

## Read in the data

```{r importgmshow}
library(data.table)
library(here)

gm <- fread(here("data", "gapminder.csv"))

# Some modifications which will be useful later
gm[, continent:=as.factor(continent)]
gm[, country:=as.factor(country)]

gm

str(gm)
```

# Life Expectancy - Portugal

Let's plot life expectancy over time for Portugal.
```{r lifeexpPT}
library(ggplot2)
library(magrittr)

gm[country == "Portugal"] %>% 
ggplot(aes(year, lifeexp)) +
  geom_point() + 
  geom_line() +
  scale_y_continuous( 
                limits=c(0, 80)) 
```


## Exercise 1: Portugal

### Infant Mortality - Portugal

Plot the infant mortality rate over time for Portugal. Include the following:

+ Colour for the lines and plots 
+ Labels for the axes.
+ Add "Infant Mortality - Portugal" as a title (hint: use `ggtitle()`)

```{r infmortPT-1}
# Type answer here.

```

**Interpret this plot.**

### Fertility Rate - Portugal

Plot the fertility rate over time for Portugal. Include the following:

+ Colour for the lines and plots 
+ Labels for the axes.
+ Add "Fertility Rate - Portugal" as a title (hint: use `ggtitle()`)

```{r fertilityPT}
# Type answer here.

```

Use the colours of the Portuguese flag,: "darkred" for points and "darkgreen" 
for lines. What do you notice when you switch the order of `geom_point()` and
`geom_line()`? Do you strictly need to include the `geom_point()`?

**Interpret this plot.**

### Population & GDP

Produce the same plots as above for population and GDP. How do you interpret them both?



# Life Expectancy - Europe

## Exercise 2: Europe

In this exercise we will evaluate how Portugal performs in Europe for:

- Infant Mortality
- Fertility Rate
- GDP per Capita

Produce the follow plots:

- Boxplot over time
- Line plot over time with European countries in grey and Portugal in blue.

### Infant Mortality - Portugal in Europe

The performance of Portugal in Europe for infant mortality rate (per 1000 births).

```{r infmort-boxplot}
# Type answer here.

```

```{r infmort-lineplot}
# Type answer here.

```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

### Fertility Rate - Portugal in Europe

The performance of Portugal in Europe for the total fertility rate (children per woman).

```{r fertility-boxplot}
# Type answer here.

```

```{r fertility-lineplot}
# Type answer here.

```


**Interpret this plot.** How does Portugal perform relative to other EU countries?

### GDP - Portugal in Europe

The performance of Portugal in Europe for the GDP per capita (inflation adjusted US$).

```{r GDP-boxplot}
# Type answer here.

```

```{r GDP-lineplot}
# Type answer here.

```


**Interpret this plot.** How does Portugal perform relative to other EU countries?

# Life Expectancy - Europe

## Exercise 3: Gapminder plot

Reproduce the "Health versus Wealth" gapminder plot but replacing health with:

- Infant Mortality
- Fertility Rate

### Gapminder Plot - Infant Mortality

```{r infmort-gdp-gapminder}
# Type answer here.

```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

### Gapminder Plot - Fertility


```{r fertility-gdp-gapminder}
# Type answer here.

```

**Interpret this plot.** How does Portugal perform relative to other EU countries?

[^gmData]: Source: https://www.gapminder.org/data/
